---
title: "Automated MCMC comparison"
subtitle: "Valencia International Bayesian Analysis Summer School Workshop"
output:
  html_document:
    toc: yes
    mathjax: default
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
## library(mcmcplots)
read_chunk('chunks_litters.R')
source("chunks_litters.R")
```

# Automating MCMC comparisons in NIMBLE

- NIMBLE provides tools to automatically compare MCMC efficiency
across methods.
- Methods can include different NIMBLE MCMC configurations as well as
  JAGS, OpenBUGS/WinBUGS (not tested much) and Stan (separate code
  required).
- We will use `compareMCMCs()`, which calls `MCMCsuite()` internally.
- `MCMCsuite()` can be used directly as well.

# Example: Metropolis-Hastings vs slice sampling. NIMBLE vs. JAGS.

 - `compareMCMCs` and `MCMCsuite` need to build and compile the model themselves.
     - (This may become more flexible in the future.)
 - These functions provide a variety of pre-defined MCMC configurations and a system
for providing your own.
 - I will illustrate using slice samplers instead of the default
Metropolis-Hastings samplers for `a` and `b` in the litters model.

```{r, prep, echo=FALSE}
# so attendees can run code below this without using code from other modules
# if(!exists('littersModel') || !exists('cLittersModels')) source('chunks_litters.R')
```                   

Here's how we set things up to provide a customized MCMC configuration as one of the comparisons.

```{r, comparisons}
littersInfo <- list(code = littersCode,
                   constants = littersConsts,
                   data = littersData, inits = littersInits)

## A configuration list for compareMCMCs
MCMCdefs <- list(
    nimble_slice = quote({
        ## This code will be evaluated in an environment
        ## where 'Rmodel' is the model.
        nodes <- c('a[1]','a[2]','b[1]','b[2]')
        MCMCconf <- configureMCMC(Rmodel) ## get a default configuration
        ## print default samplers for illustration
        MCMCconf$printSamplers(nodes)
        for(node in nodes) {
                 MCMCconf$removeSamplers(node)
                 MCMCconf$addSampler(target = node, type = "slice")
        }
        ## print modified samplers for illustration
        MCMCconf$printSamplers(nodes)
        MCMCconf
    })
)
```

# Running MCMCsuite and compareMCMCs

You can either use `MCMCsuite()` or `compareMCMCs()`. The latter allows you to print out some nice html plots of the comparisons.

(I'm not showing the output here as the functions are not really designed to interface with compiling an R Markdown document.)

```{r, compareMCMC, eval=FALSE}
littersComparisons <- compareMCMCs(
    littersInfo,
    MCMCs = c("jags", "nimble", "nimble_slice"),
    MCMCdefs = MCMCdefs,
    monitors = c("a", "b"),
    niter = 10000,
    burnin = 1000,
    summary = FALSE)

## or here is the MCMCsuite call
littersResults <- MCMCsuite(
               littersCode,
               constants = littersConsts,
               data = littersData,
               inits = littersInits,
               MCMCs = c("jags", "nimble", "nimble_slice"),
               MCMCdefs = MCMCdefs,
               monitors = c("a", "b"),
               niter = 10000,
               burnin = 1000)
```

# Viewing the comparisons

```{r, comparison-results, eval=FALSE}
make_MCMC_comparison_pages(littersComparisons,
                           dir = "litters_comparison_results",
                           modelNames = "littersComparisons")
```

The results are [here](litters_comparison_results/littersComparisons.html).

