---
title: "How NIMBLE's model language differs from BUGS and JAGS"
subtitle: "Valencia International Bayesian Analysis Summer School Workshop"
output: html_document
---

```{r setup, include=FALSE} 
library(methods) ## needed only when building documents outside of R
library(nimble)
source("chunks_litters.R")
```

# How NIMBLE is the same as BUGS and JAGS

* Most distributions and functions are supported

    - For distributions, see <a href="http://r-nimble.org/manuals/NimbleUserManual.pdf#page=39" target="_blank" style="color: blue">User Manual Section 5.2.4</a>
    - For functions, see  <a href="http://r-nimble.org/manuals/NimbleUserManual.pdf#page=44" target="_blank" style="color: blue">User Manual Section 5.2.5</a>
	
* Most syntax is supported

    - Truncation syntax is different when using `nimbleCode`.  (It can
      be the same as for JAGS if reading code from a file with `readBUGSmodel()`).

# How NIMBLE extends BUGS

  - Alternative distribution parameterizations (like R).
  - Named parameters (like R).
  - Vectorized math and linear algebra.
  - Definition-time if-then-else (multiple model variants from the same code).
  - User-defined functions and distributions.

### How NIMBLE is limited compared to BUGS and/or JAGS:

   - NIMBLE is stricter about requiring square brackets and
   informative indices for non-scalar nodes.

# R-like alternative and named parameters

Note named parameters in distributions.

```{r, litters-model-param, eval=FALSE}

littersCode <- nimbleCode({
  for (i in 1:G) {
     for (j in 1:N) {
     	# likelihood (data model)
        r[i,j] ~ dbin(p[i,j], n[i,j])
	# latent process (random effects)
        p[i,j] ~ dbeta(a[i], b[i]) 
     }
     # prior for hyperparameters
     a[i] ~ dgamma(shape = 1, rate = .001)
     b[i] ~ dgamma(shape = 1, scale = 1/.001)
   }
})
```

Note that we could have used the mean/sd parameterization of the beta distribution, which would then require different hyperparameter specification.

# Extensibility: new functions and distributions

We will cover this in depth later.

# Distinction between data and constants

- Constants must be provided to `nimbleModel`

    - Constants are values needed to define model relationships

        - Index ranges like *N* in the litters model
        - Constant vectors used for indexing: e.g., *block* in `mu[block[i]]`

- Data can be provided to `nimbleModel` or later

    - Data represents a flag on the role graph nodes play in the model

        - Data nodes shouldn't be sampled in an MCMC

    - Data values can be changed.

Here's an example:

```{r, prep, echo=FALSE}
# so attendees can run code below this without using code from other modules
# if(!exists('littersModel') || !exists('cLittersModels')) source('chunks_litters.R')
```                   


```{r, data-example}
littersModel$isData('r')
littersModel$isData('p')
littersModel$r
littersModel$p
littersModel$simulate('r')
littersModel$simulate('p')
littersModel$r
littersModel$p
littersModel$simulate('r', includeData = TRUE)
littersModel$r
```

- NIMBLE will usually disambiguate data when it is provided as constants.

# Constants vs. right-hand side nodes

Regression predictors (i.e., covariates or independent variables) can be treated either as constants or as variables in the model. The latter allows one to change the values after the model is built. 

When a variable is specified only on the right-hand side of expression(s) in the model, it does not become a graph node in the model because it is not a parameter or data, but it is a variable in the model that can be assigned new values. 

(Regression predictors can also be specified as *data* though they do not play the role of data in terms of the likelihood.)

# More explicit need to provide dimensions (in models, not in nimbleFunctions)

- Square brackets must always be provided to indicate number of dimensions

    - If `x` is 2-dimensional, use `x[,] %*% beta[]`, not `x %*% beta`
    
* Sometimes NIMBLE is not as smart as BUGS/JAGS at determining dimensions.  There are two solutions:

    * Give dimensions explicity: `x[1:n, 1:m]`, OR
    * Provide a `dimensions` argument to `nimbleModel`.  Example: `dimensions = list(x = c(n, m))`.

# Vectorized math and linear algebra

Instead of writing this in your model code:

```{r, vectorized-1}
nimbleOptions(verbose = FALSE)
m1 <- nimbleModel(
    nimbleCode({
        for(i in 1:5) {
            predicted[i] <- beta0 + beta1 * x[i]
        }
    }
    ))
```
you can write this:
```{r, vectorized-2}
m2 <- nimbleModel(
    nimbleCode({
        predicted[1:5] <- beta0 + beta1 * x[1:5]
    }
    ))
```

These are not equivalent:
```{r, vectorized-3}
## m1 has 5 scalar nodes
m1$getNodeNames()
## m2 has 1 vector node
m2$getNodeNames()
```

One is not necessarily better than the other.  It depends on the model
and the MCMC configuration.  (More on those topics later.)

Vectorized declarations do not work for univariate distributions (e.g., Normal, beta, or gamma distributions)

However, vectorized declarations are necessary for declaring multivariate distributions


# Definition-time if-then-else

If you wish to define multiple alternative models in one set of code,
you can use if-then-else statements.  These will be evaluated based on
variables in the R environment when the model is defined.  For
example:

```{r if-then-else}
code <- nimbleCode({
    sigma ~ dunif(0, 10)
    beta0 ~ dnorm(0, sd = 1000)
    beta1 ~ dnorm(0, sd = 1000)
    if(INCLUDE_X2) { beta2 ~ dnorm(0, sd = 1000) } else {}
    for(i in 1:10) {
        if(INCLUDE_X2) {
            y[i] ~ dnorm(beta0 + beta1 * x1[i] + beta2 * x2[i], sd = sigma)
        } else {
            y[i] ~ dnorm(beta0 + beta1 * x1[i], sd = sigma)
        }
    }
})

INCLUDE_X2 <- FALSE
m1 <- nimbleModel(code)
INCLUDE_X2 <- TRUE
m2 <- nimbleModel(code)
m1$getNodeNames()
m2$getNodeNames()

```

m2 has `beta2` while m1 does not.  The long names are "lifted nodes" -- more on those later.

