---
title: "Exercises: Writing nimbleFunctions"
subtitle: "Valencia International Bayesian Analysis Summer School Workshop"
output: html_document
---

# Exercise 1


Let's consider using a nimbleFunction to replace a for loop that can't be avoided in R. Write a second order random walk using a nimbleFunction. Here's the code for the R version. 

```{r, markov-exer, eval=FALSE}
set.seed(0)
n <- 1e6
path <- rep(0, n)
rho1 <- .8
rho2 <- .1
path[1:2] <- rnorm(2)
print(system.time(
for(i in 3:n)
      path[i] <- rho1*path[i-1] + rho2*path[i-2] + rnorm(1)
))
ts.plot(path[1:5000])
```

Now fill out the nimbleFunction version and test the timing.

```{r, markov-exer-scaffold, eval=FALSE}
mc <- nimbleFunction(
   run = function( ... ) ) {
       returnType( ... )
       ...
       return(...)
})
cmc <- compileNimble(mc)
set.seed(1)
system.time(path <- cmc(n, rho1, rho2))
```

Extra:  Use *nimStop()* as part of an error check that ensures that the length of the path to be sampled is longer than the order of the dependence. 

# Exercise 2

2) Write a nimbleFunction that will simulate from the prior for an arbitrary model $m$ times and store the results in a modelValues object. (E.g., this could be used as part of a simulation study.) Time how long this takes running both the uncompiled and compiled nimbleFunction on a model of your choice. (Feel free to use the litters model.)
