---
title: "nimbleFunctions: Writing MCMC samplers"
subtitle: "Valencia International Bayesian Analysis Summer School Workshop"
output: html_document
---


```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found - weird
library(nimble)
```

# Introduction

NIMBLE's MCMC system was designed to be extensible - it's easy for users to add samplers and use them right away for their models.

First we'll see how nimbleFunctions are used to write NIMBLE's built-in samplers.
Then in the next module we'll see how one can write one own's sampler and use it in your MCMC, along with all of NIMBLE's built-in samplers.


# NIMBLE's built-in random walk sampler

Here's a simplified version of the random walk sampler used for NIMBLE's MCMC engine. Note that this simplification excludes the use of an adaptive algorithm that adjusts the proposal standard deviation to the acceptance rate as the MCMC iterations proceed.

```{r, basic-sampler, eval=FALSE}
sampler_RW <- nimbleFunction(
    contains = sampler_BASE,  # some object-oriented syntax for MCMC samplers
    setup = function(model, mvSaved, target, control) {
        # All samplers must take the above arguments to work with NIMBLE's MCMC engine
        targetAsScalar <- model$expandNodeNames(target, 
                                                returnScalarComponents = TRUE)
        if(length(targetAsScalar) > 1)   
           stop('cannot use RW sampler on more than one target; try RW_block sampler')
        if(model$isDiscrete(target))     
           stop('cannot use RW sampler on discrete-valued target; try slice sampler')
        ###  control list extraction  ###
        scale <- control$scale
        ###  node list generation  ###
        calcNodes  <- model$getDependencies(target)
    },

    run = function() {
        ## this assumes logProb values are up-to-date,
        ## which is how all samplers should be set up
        logProbCurrent <- model$getLogProb(calcNodes)
        ## 'target', 'model' is available because it was a
        ## setup code argument
        currentValue <- model[[target]]
        # 'scale', 'calcNodes' available because created in setup code
        propValue <- rnorm(1, mean = currentValue, sd = scale)
        model[[target]] <<- propValue
        model$calculate(calcNodes)
        logMHR <- model$getLogProb(calcNodes) - logProbCurrent
        jump <- decide(logMHR)
        ## make sure state of model is up-to-date and mvSaved
        ## has current iteration
        if(jump) nimCopy(from = model, to = mvSaved, row = 1, 
                         nodes = calcNodes, logProb = TRUE)
        else     nimCopy(from = mvSaved, to = model, row = 1, 
                         nodes = calcNodes, logProb = TRUE)
    }
)
```


# NIMBLE's Metropolis sampler


Let's look at the code for the NIMBLE's full Metropolis sampler. You can find this in the file *R/MCMC_samplers.R* in the source code for the NIMBLE package (note that ```nimble:::sampler_RW``` is not useful to look at, as it shows a modified version of this, namely the nimbleFunction after it has been built).

```{r, Metr}
sampler_RW <- nimbleFunction(
    contains = sampler_BASE,
    setup = function(model, mvSaved, target, control) {
        ## control list extraction
        logScale      <- if(!is.null(control$log))           control$log           else FALSE
        reflective    <- if(!is.null(control$reflective))    control$reflective    else FALSE
        adaptive      <- if(!is.null(control$adaptive))      control$adaptive      else TRUE
        adaptInterval <- if(!is.null(control$adaptInterval)) control$adaptInterval else 200
        scale         <- if(!is.null(control$scale))         control$scale         else 1
        ## node list generation
        targetAsScalar <- model$expandNodeNames(target, returnScalarComponents = TRUE)
        calcNodes <- model$getDependencies(target)
        ## numeric value generation
        scaleOriginal <- scale
        timesRan      <- 0
        timesAccepted <- 0
        timesAdapted  <- 0
        scaleHistory  <- c(0, 0)
        acceptanceHistory  <- c(0, 0)
        if(nimbleOptions('saveMCMChistory')) {
            saveMCMChistory <- TRUE
        } else saveMCMChistory <- FALSE
        optimalAR     <- 0.44
        gamma1        <- 0
        ## checks
        if(length(targetAsScalar) > 1)   stop('cannot use RW sampler on more than one target; try RW_block sampler')
        if(model$isDiscrete(target))     stop('cannot use RW sampler on discrete-valued target; try slice sampler')
    },
    
    run = function() {
        currentValue <- model[[target]]
        propLogScale <- 0
        if(logScale) { propLogScale <- rnorm(1, mean = 0, sd = scale)
                       propValue <- currentValue * exp(propLogScale)
                   } else         propValue <- rnorm(1, mean = currentValue,  sd = scale)
        model[[target]] <<- propValue
        logMHR <- calculateDiff(model, calcNodes) + propLogScale
        jump <- decide(logMHR)
        if(jump) nimCopy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
        else     nimCopy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
        ## call additional method (defined below)
        if(adaptive)     adaptiveProcedure(jump)
    },

    ## additional methods, callable by nimbleFunction$method_name,
    ## or by method_name within a given nimbleFunction's run function or methods
    methods = list(
        
        adaptiveProcedure = function(jump = logical()) {
            timesRan <<- timesRan + 1
            if(jump)     timesAccepted <<- timesAccepted + 1
            if(timesRan %% adaptInterval == 0) {
                acceptanceRate <- timesAccepted / timesRan
                timesAdapted <<- timesAdapted + 1
                if(saveMCMChistory) {
                    setSize(scaleHistory, timesAdapted)         
                    scaleHistory[timesAdapted] <<- scale        
                    setSize(acceptanceHistory, timesAdapted)    
                    acceptanceHistory[timesAdapted] <<- acceptanceRate  
                }
                # Shaby & Wells (2011) adaptation scheme
                gamma1 <<- 1/((timesAdapted + 3)^0.8)
                gamma2 <- 10 * gamma1
                adaptFactor <- exp(gamma2 * (acceptanceRate - optimalAR))
                scale <<- scale * adaptFactor
                timesRan <<- 0
                timesAccepted <<- 0
            }
        },
        
        getScaleHistory = function() {
            returnType(double(1))
            if(saveMCMChistory) {
                return(scaleHistory)
            } else {
                print("Please set 'nimbleOptions(saveMCMChistory = TRUE)' before building the MCMC")
                return(numeric(1, 0))
            }
        },          

        getAcceptanceHistory = function() {
            returnType(double(1))
            if(saveMCMChistory) {
                return(acceptanceHistory)
            } else {
                print("Please set 'nimbleOptions(saveMCMChistory = TRUE)' before building the MCMC")
                return(numeric(1, 0))
            }
        },          

        reset = function() {
            scale <<- scaleOriginal
            timesRan      <<- 0
            timesAccepted <<- 0
            timesAdapted  <<- 0
            if(saveMCMChistory) {
                scaleHistory  <<- c(0, 0)
                acceptanceHistory  <<- c(0, 0)
            }
            gamma1 <<- 0
        }
    )
)

```

Much of that code has to do with:

  - making the sampler adaptive, so that the proposal scale adapts so that a good acceptance rate is achieved, and
  - saving information about sampler behavior.

# nimbleFunctions and object-oriented programming

Notice the class structure of a nimbleFunction:

   - various methods, in addition to the *run* function
   - member 'data' within the class that can be assigned to with `<<-`
   - methods and member 'data' can be called/accessed directly by name
   - setup code literally does produce a class (an R reference class, which becomes a C++ class in compiled code)
