---
title: "Bayesian nonparametrics example (1): Meta analysis"
subtitle: "Valencia International Bayesian Analysis Summer School Workshop"
output: html_document
---


```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found 
library(nimble)
```

# Introduction

Here we'll consider a real example, doing a meta-analysis. We'll start with a standard random effects meta analysis and then robustify the analysis using Bayesian nonparametric methods.



# Meta analysis example

Meta analysis seeks to combine results across multiple studies of the same phenomenon to increase power. It's often applied to clinical trials.

The example is of the side effects of a very popular drug for diabetes called Avandia. The question is whether Avandia use increases the risk of myocardial infraction (heart attack). There are 48 studies (the 49th study in the data file is different in some ways and excluded here), each with treatment and control arms.

```{r, avandia-view}
dat <- read.csv('avandia.csv')
head(dat)
```

Here we'll start with a generalized linear mixed model (GLMM)-based meta analysis. In fact the model is not so different than our example litters model.

# Basic meta analysis of Avandia MIs

```{r, avandia-setup}
dat <- read.csv('avandia.csv')
dat <- dat[-49, ]

x <- dat$controlMI
n <- dat$nControl
y <- dat$avandiaMI
m <- dat$nAvandia

nStudies <- nrow(dat)
data <- list(x = x, y = y)
constants = list(n = n, m = m, nStudies = nStudies)

codeParam <- nimbleCode({
    for(i in 1:nStudies) {
        y[i] ~ dbin(size = m[i], prob = q[i]) # avandia MIs
        x[i] ~ dbin(size = n[i], prob = p[i]) # control MIs
        q[i] <- expit(theta + gamma[i])       # Avandia log-odds
        p[i] <- expit(gamma[i])               # control log-odds
        gamma[i] ~ dnorm(mu, sd = tau)        # study effects
    }
    theta ~ dflat()        # effect of Avandia
    # random effects hyperparameters
    mu ~ dflat()
    tau ~ dunif(0, 100)
})
```

$\theta$ quantifies the difference in risk between the control and treatment arms, while the $\gamma_i$ quantify study-specific variation using normally-distributed random effects.

# Running the MCMC

Let's run a basic MCMC.

```{r, mcmc, fig.cap='', fig.width=12, fig.height=5}
inits = list(theta = 0, mu = 0, tau = 1, gamma = rnorm(nStudies))

samples <- nimbleMCMC(code = codeParam, data = data, inits = inits,
                      constants = constants, monitors = c("mu", "tau", "theta", "gamma"),
                      thin = 10, niter = 21000, nburnin = 1000, nchains = 1, setSeed = TRUE)
gammaCols <- grep('gamma', colnames(samples))

par(mfrow = c(1, 4))
ts.plot(samples[ , 'theta'], xlab = 'iteration', ylab = expression(theta))
hist(samples[ , 'theta'], xlab = expression(theta), main = 'effect of Avandia')
gammaMn <- colMeans(samples[ , gammaCols])
hist(gammaMn, xlab = 'posterior means of random effects', main = 'random effects distribution')
hist(samples[1000, gammaCols], xlab = 'single draw of random effects',
                   main = 'random effects distribution')
```

What about the normality assumption - could the estimated distributions be skewed (and remember that these are generated under the normality assumption).


# DP-based random effects modeling for meta analysis

```{r, meta-bnp}
codeBNP <- nimbleCode({
    for(i in 1:nStudies) {
        y[i] ~ dbin(size = m[i], prob = q[i]) # avandia MIs
        x[i] ~ dbin(size = n[i], prob = p[i]) # control MIs
        q[i] <- expit(theta + gamma[i])       # Avandia log-odds
        p[i] <- expit(gamma[i])               # control log-odds
        gamma[i] ~ dnorm(muTilde[xi[i]],      # random effects (from mixture)
                     var = tauTilde[xi[i]])   # with mean/var from one component of mixture
    }
    # mixture component parameters drawn from base measures
    for(i in 1:nStudies) {
        muTilde[i] ~ dnorm(mu0, sd = sd0)
        tauTilde[i] ~ dinvgamma(a0, b0)
    }
    # CRP for clustering studies to mixture components
    xi[1:nStudies] ~ dCRP(alpha, size = nStudies)
    # hyperparameters
    alpha ~ dgamma(1, 1)      
    mu0 ~ dflat()
    sd0 ~ dunif(0, 100)
    a0 ~ dunif(0, 100)
    b0 ~ dunif(0, 100)
    theta ~ dflat()          # effect of Avandia
})
```

The specification is a bit complicated, but just think of it as using a nonparametric extension to a mixture of normal distributions as the random effects distribution for $\gamma_i$, but where we don't fix the maximum number of components.

# Running an MCMC for the DP-based meta analysis

```{r, DP-MCMC, fig.cap='', fig.width=12, fig.height=5}
inits <- list(gamma = rnorm(nStudies), xi = sample(1:2, nStudies, replace = TRUE),
              alpha = 1, mu0 = 0, sd0 = 1, a0 = 1, b0 = 1, theta = 0,
              muTilde = rnorm(nStudies), tauTilde = rep(1, nStudies))

samplesBNP <- nimbleMCMC(code = codeBNP, data = data, inits = inits,
               constants = constants,
               monitors = c("theta", "gamma", "alpha", "xi", "mu0", "sd0", "a0", "b0"),
               thin = 10, niter = 21000, nburnin = 1000, nchains = 1, setSeed = TRUE)

gammaCols <- grep('gamma', colnames(samplesBNP))
xiCols <- grep('xi', colnames(samplesBNP))

par(mfrow = c(1,5))
ts.plot(samplesBNP[ , 'theta'], xlab = 'iteration', ylab = expression(theta))
hist(samplesBNP[ , 'theta'], xlab = expression(theta), main = 'effect of Avandia')
gammaMn <- colMeans(samplesBNP[ , gammaCols])
hist(gammaMn, xlab = 'posterior means of random effects',
              main = 'random effects distribution')
hist(samplesBNP[1000, gammaCols], xlab = 'single draw of random effects',
                   main = 'random effects distribution')

# How many mixture components are inferred?
xiRes <- samplesBNP[ , xiCols]
nGrps <- apply(xiRes, 1, function(x) length(unique(x)))
ts.plot(nGrps, xlab = 'iteration', ylab = 'number of components')
```

Conclusions: the primary inference seems robust, and there's also not much evidence of multiple components.

What samplers are being used? `nimbleMCMC` doesn't tell us, but we could configure the default MCMC to see:

```{r, DP-samplers}
model <- nimbleModel(codeBNP, constants = constants, data = data, inits = inits)
conf = configureMCMC(model, print = TRUE)
```
